---
title: "Visualizing large single-cell and spatial datasets with MANDALA"
author: "Nicolas J. Ho, Arvind S. Mer"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_width: 7
    fig_height: 5.5
vignette: >
  %\VignetteIndexEntry{Visualizing large single-cell and spatial datasets with MANDALA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  dpi = 96,
  fig.align = "center"
)
```

```{r libs}
library(MANDALA)
library(Seurat)
```

# Why aggregate

A UMAP scatter plot spends its only spare encoding channel, colour, on a
single metadata variable such as cell types, and loses local density to 
overplotting. Past roughly $10^5$ cells the rendered image depends on the 
order in which rows happen to be drawn, with cells drawn later masking 
previously drawn cells. Populations that are rare relative to their neighbours 
therefore disappear beneath abundant ones at any practical figure size.

MANDALA aggregates before rendering. Cells are pooled into bubbles or bins,
attributes are summarised within each aggregated unit, and the summaries are 
mapped onto distinct visual channels depending on the function being called. 
Output is deterministic with respect to row order because aggregation precedes 
drawing, and the number of graphical objects is set by the aggregation 
resolution rather than by the cell count.

| Function | Output | Purpose |
| --- | --- | --- |
| `MANDALA` | $Shiny, interactive$ | $Interactive point-and-click access to all functions below$ |
| `BubbleMAP3D` | interactive HTML | The same representation with an expression height axis |
| `SpatialMAP3D` | interactive HTML | The same viewer over tissue coordinates |
| `BubbleMAP` | static (ggplot) | Density contours with data-driven bubble aggregation |
| `HexMAP` | static (ggplot) | Hexagonal binning: majority, entropy, mean expression |
| `MagnicMAP` | static (ggplot) | Paired overview and magnified views |


# Example data

```{r data}
data(sample_scRNA)     # 8 marker genes, 11 annotated cell types, UMAP precomputed
data(sample_spatial)   # bone marrow Visium: 4 genes, 1,714 spots, slice1 image

sample_scRNA
table(sample_scRNA$cell_type)
```

`sample_scRNA` is a feature-reduced version containing all of the original 
391,963 cells, it is a pan-cancer tumour microenvironment atlas (Rapozo 
Guimaraes et al., 2024), from CZ CELLxGENE Discover.

It has a `counts` layer for exploring the package. This is a property of
the bundled example, not a general recommendation: for your own objects, use the
normalised layer.

`sample_spatial` is a visium spatial object from an Acute Myeloid Leukemia bone
marrow biopsy (Enes Dasdemir et al., 2025 ). Metadata such as RCTD deconvolution 
labels and tissue structures have been added in `meta.data`, any of which can be 
placed on the height axis with `metric =`.

The most straight-forward way to start using the package is by MANDALA
using:

```{r app-quickstart, eval = FALSE}
MANDALA(sample_scRNA) #and select BubbleMAP3D, or other scRNA-seq 
			   #visualizing function, in the pop up or

MANDALA(sample_spatial) #and select SpatialMAP3D in the popup
```

# BubbleMAP

## How bubbles are formed

A greedy pass repeatedly picks an unassigned seed cell, absorbs every unassigned
cell of the same class within `radius`, and emits one bubble whose area encodes
the number of cells absorbed. Because bubbles form from the cells themselves
rather than from a fixed grid, their positions follow the local structure of the
embedding rather than an arbitrary lattice.

Neighbour queries use a uniform spatial hash whose cell size equals `radius`.
Any point within `radius` of a seed must lie in the seed's own grid cell or one
of the eight adjacent ones, so scanning that 3×3 block and applying the distance
test is exact and returns the same neighbours a full radius search would,
without ever materialising the complete neighbour list.

## Count mode

```{r bubble-count, fig.height = 6}
BubbleMAP(
  sample_scRNA,
  reduction       = "umap",
  label           = "cell_type",
  radius          = 1,
  min_group       = 10,
  min_class       = 150,
  target_cells    = NULL,
  bubble_min_dens = NULL,
  size_by         = "count",
  celltype_legend = TRUE,
  label_size      = 3
)
```

## Gene mode

Bubble area still encodes cell count; shape distinguishes bubbles whose mean
expression falls above or below `expr_cutoff`. Keeping area on the count means
sizes stay comparable across panels whatever else is being encoded.

```{r bubble-gene, fig.height = 6}
BubbleMAP(
  sample_scRNA,
  reduction    = "umap",
  label        = "cell_type",
  radius       = 1,
  min_group    = 10,
  min_class    = 150,
  target_cells = NULL,
  size_by      = "gene",
  gene         = "CLDN5",
  assay        = "RNA",
  slot         = "counts",
  expr_cutoff  = 0.5,
  highlight    = "endothelial cell",
  celltype_legend = TRUE
)
```

A threshold encoding suits markers that are effectively binary. For a marker
detected in only a few percent of cells, shape works where a continuous
magnitude scale would not: there is no range to spread across.

## Metric mode

A continuous metadata column, encoded either as bubble transparency (default) or
as a viridis fill with the cell-type colour retained on the bubble outline.

```{r bubble-metric, fig.height = 6}
BubbleMAP(
  sample_scRNA,
  reduction    = "umap",
  label        = "cell_type",
  radius       = 1,
  min_group    = 10,
  size_by      = "metric",
  metric       = "percent.mt",
  metric_scale = "colour",   # "alpha" for the transparency encoding
  metric_edge  = TRUE,       # ring each bubble by cell type
  celltype_legend = TRUE
)
```

`metric_scale = "colour"` is usually the better choice: alpha is a weak
perceptual channel, and the density contours underneath flatten it further.
With `metric_edge = TRUE` the class identity is still readable per bubble,
because it moves to the outline.

## Key parameters

| Parameter | Default | Notes |
| --- | --- | --- |
| `radius` | `0.30` | Neighbourhood radius. Larger values merge more aggressively; check bubble counts before settling. |
| `min_group` | `8` | Minimum cells per bubble. |
| `min_class` | `200` | Populations below this are pooled into "Other" and dropped from labels and legend. |
| `rare_relax` | `3` | Classes with fewer than `rare_relax * min_group` cells get a proportionally reduced threshold. `NULL` disables. |
| `target_cells` | `120000` | Stratified per-class cap. `NULL` uses every cell. |
| `bubble_min_dens` | `0.10` | Drops bubbles outside their own class density envelope. `NULL` keeps all. |
| `highlight` | `NULL` | Class names to emphasise. Must match the metadata exactly. |
| `bubble_engine` | `"hash"` | Spatial-hash neighbour lookup. `"legacy"` uses RANN. |

`min_class` is the parameter most likely to surprise you. If a cell type
disappears from the plot, check its size against `min_class` before adjusting
anything else. Classes that form no bubbles are reported by name when the
function runs.

`radius` and `min_group` interact: a bubble needs `min_group` cells within
`radius`, so raising the radius and the threshold together keeps bubble counts
roughly stable while changing how much local structure is smoothed away.

## Return value

```{r bubble-return}
res <- BubbleMAP(sample_scRNA, label = "cell_type", layer = "counts")

names(res)
nrow(res$bubbles)
head(res$bubbles)
tapply(res$bubbles$count, res$bubbles$class, sum)
```

`bubbles` has one row per bubble, with `count` and any `expr_value` /
`metric_value`. Also returned: `plot`, `contours`, `peaks`, `palette` and
`class_n`. Pass `keep_data = TRUE` to additionally get the per-cell data frame,
which is omitted by default because it keeps the full object alive after
plotting.

# BubbleMAP3D

The same bubble reduction, written to a self-contained HTML viewer. Each bubble
is lifted to the mean expression of a chosen marker or a metadata column of some 
scores, so cell types that overlap in the plane separate by height.

```{r live, eval = FALSE}
BubbleMAP3D(
  sample_scRNA,
  gene   = c("CLDN5", "COL1A1", "KRT14", "S100A8",
             "NKG7", "LYZ", "IGKC", "TCL1A"),
  metric = c("percent.mt"),          # metadata columns work as height too
  label  = "cell_type",
  layer  = "counts"
)
```

All markers are computed at once, so switching between them in the viewer is
instant. In the viewer: drag to rotate, scroll to zoom, click a legend swatch to
recolour a type or a row to hide it, and export vector PDF or 300 dpi PNG at a
chosen figure width.

## Why a bubble mean makes sparse markers usable

A marker detected in 1% of cells is far too zero-inflated for a per-cell height
axis: almost every cell sits on the floor, and the surface collapses to a flat
sheet with a few isolated spikes. Averaged over the roughly 150 cells in a
bubble, the same marker has real spread and produces legible relief.

This is the main reason the height axis operates on bubbles rather than cells.
It also means the height of a bubble is a local average, not a cell-level
measurement, i.e. a bubble is raised because its neighbourhood expresses, not
because any particular cell does.

## Useful arguments

- `flat = TRUE` opens overhead with height off, giving the 2-D layout.
- `shell` draws a contour envelope around the raised type.
- `interactive_cells` controls how many cells are shipped for the viewer's live
  regrouping sliders; `NULL` omits them and keeps the file small.
- `elevation`, `highlight_all` and `highlight_lifted` set the opening state of
  the controls.


# SpatialMAP3D

The same viewer with the floor taken from tissue coordinates rather than a
reduction. Coordinates come from a Seurat image or field-of-view (FOV), or 
from named `meta.data` columns via `coords =`.

```{r spatial, eval = FALSE}
SpatialMAP3D(
  sample_spatial,
  gene      = c("COL1A1", "HBB", "MPO", "NKG7"),
  metric    = c("dist_to_adipo"), 
  image     = "slice1",
  label     = "MQ_compartment",
  mode      = "point",
  assay     = "SCT",
  layer     = "data",
  min_class = 20
)
```

`mode = "point"` draws one mark per spot, which is usually right for Visium and
for segmented imaging data: those are already the unit of measurement, so
pooling them discards resolution. `mode = "bubble"` pools them, which is what
you want for dense single-cell spatial. The viewer's "Original points" button
switches between the two.

Coordinates are rescaled internally so that every length in the viewer behaves
the same whether the data is in microns, pixels or array indices. `radius` is
given in your original units and converted for you; if omitted it is derived
from the measured spot spacing.


# HexMAP

Bins cells into a hexagonal lattice and summarises each bin. `hex_density` sets
the number of bins across the shorter axis; bins containing fewer than
`min_cells` cells are dropped.

## Majority composition

```{r hex-majority, fig.height = 5.5}
HexMAP(sample_scRNA, reduction = "umap", label_col = "cell_type",
        hex_density = 70, plot_type = "majority", min_cells = 3,
        alpha = 0.3)
```

Alpha is scaled to the majority proportion, so identity and its local purity are
read together: a washed-out bin is a mixed one.

## Shannon entropy

```{r hex-entropy, fig.height = 5.5}
HexMAP(sample_scRNA, reduction = "umap", label_col = "cell_type",
        hex_density = 70, plot_type = "entropy", min_cells = 3,
        viridis = "magma")
```

Entropy makes transition zones and annotation ambiguity directly visible rather
than inferred. High-entropy regions are where cell-type calls disagree most.

## Mean expression

```{r hex-gene, fig.height = 5.5}
HexMAP(sample_scRNA, reduction = "umap", label_col = "cell_type",
        hex_density = 70, plot_type = "gene", gene_name = "CLDN5",
        assay = "RNA", layer = "counts", min_cells = 3)
```

Hexagonal binning of single-cell embeddings was introduced by
[schex](https://bioconductor.org/packages/schex) (Freytag and Lister, 2020).
Users who need binning alone should consider it directly; `HexMAP` is provided
so that binning shares an interface and a palette with the other methods here.

# MagnicMAP

Draws an overview panel with the region of interest outlined, and a magnified
panel beneath it, so the inspected region stays locatable within the whole
embedding.

```{r magnic, eval = requireNamespace("patchwork", quietly = TRUE), fig.height = 7}
MagnicMAP(
  sample_scRNA,
  point_of_interest = c(0, 5),
  reduction     = "umap",
  group.by      = "cell_type",
  zoom_factor   = 4,
  pt.size       = 0.5,
  label         = TRUE,
  label.size    = 2,
  combine_plots = TRUE
)
```

`combine_plots = FALSE` returns `main_plot`, `zoom_plot` and `zoom_coordinates`
separately, which is useful when the two panels need different point sizes.

# MANDALA

A local Shiny front end over an object already in your session. Nothing is
uploaded and there is no size limit beyond your own memory.

```{r app, eval = FALSE}
MANDALA(sample_scRNA)
```

1. Pick the assay and data slot.
2. Pick a `Grouping Column` from the metadata such as cell type columns.
3. Select `BubbleMAP3D` for scRNA-seq data, or `SpatialMAP3D` for spatialRNA-seq data
4. Select reductions/image, genes and/or metrics.
5. Render. 

Figures save to a folder you choose through a native dialog: PDF for the static
methods, HTML for the interactive viewers. Results are cached per parameter set,
and a cell cap is offered for larger objects.


# Session info

```{r session}
sessionInfo()
```

# References

Freytag, S. and Lister, R. (2020) schex avoids overplotting for large
single-cell RNA-sequencing datasets. *Bioinformatics*, **36**, 2291–2292.

Rapozo Guimaraes, G. et al. (2024) Single-cell pan-cancer tumour
microenvironment atlas. CZ CELLxGENE Discover.

Dasdemir, E. et al. (2025) Integrative spatial multi-omics reveal niche-
specific inflammatory signaling and differentiation hierarchies in AML. 
iScience, 29(1), 114289.
